Legion: Expressing Locality and Independence with Logical Regions

### 2012-05-02 doc/sc2012/soop.tex

\item For performance, Legion must hide the extremely long latencies associated
  with machines that have both distributed memory and many levels of
  memory hierarchy.
\end{itemize}
Our scheduler uses three techniques to achieve both correctness and high performance:
\begin{itemize}
\item If it were necessary to compare all pairs of tasks for dependences that alone
would be a serial bottleneck in all but the simplest Legion programs.  Fortunately,
subtasks enjoy an isolation property that allows Legion to check only task siblings for dependences
(see Section~\ref{sec:dep}).

@@

Each processor in the machine runs an instance of the SOOP, which handles all requests
for Legion services from tasks executing on the processor.  The SOOP is analagous to the out-of-order
instruction scheduler in superscalar processors, but working at the granularity of tasks and their region
arguments instead of instructions and their register arguments.
Legion's SOOP has a five stage pipeline: dependence analysis (Section~\ref{sec:dep}),
distribution (Section~\ref{sec:dist}),
mapping (Section~\ref{sec:map}),
execution (Section~\ref{sec:exec}),
and clean-up (Section~\ref{sec:clean}).
We discuss each in turn.

@@

When a Legion task runs, it may spawn new subtasks to be executed.  We refer to the
subtasks as {\em children} of the {\em parent} task.  The children of a parent task
are {\em siblings} of each other.

When a parent task spawns a child task, the child is {\em registered} with the SOOP 
on which the parent is executing; registration records the subtask's logical region
arguments and associated privileges and coherence properties.
Children are registered in the sequential execution order in which the parent task
spawns them.

If two children have a dependence between their region arguments, the
SOOP will schedule them to run serially to preserve the program's
sequential semantics.  If two child tasks have no dependence then
Legion can schedule them to execute out of order, in parallel.  Thus,
the first stage in the execution of a task is dependence analysis
between tasks.

\subsubsection{Dependences}
Conceptually, detecting dependences between a newly registered task $t$ and a previous task 
$t'$ requires comparing the two sets of logical regions accessed.  For each logical region used by
$t'$ that may overlap with a logical region used by $t$, the privileges
and coherence modes are compared to determine whether a dependence exists.  If
both regions need only read privileges there is never a dependence, but if either
task needs write or reduction privileges, the coherence modes are compared:
{\small
\begin{tabular}{c|cccc}
             & Exclusive & Atomic   & Simultaneous & Relaxed \\
\midrule
Exclusive    & Dep & Dep & Dep & Dep \\ 
Atomic       & Dep & Same & Cont & Cont \\
Simultaneous & Dep & Cont & Same & None \\
Relaxed      & Dep & Cont & None & None \\
\end{tabular}
}

A {\tt Dep} entry indicates dependence while {\tt None}
indicates independence.  A {\tt Same} entry is a dependence unless the two tasks
use the same physical instance of the logical region.\footnote{For example, if two tasks wish to acess the region with atomic coherence, then if they are working on the same physical copy of the data atomicity can be guaranteed using, e.g., locking.  If they are working on different copies of the data then the only safe execution strategy for Legion is to serialize the two tasks.}
A {\tt Cont} entry
indicates a dependence unless a single writer is using the
atomic coherence mode.  When a dependence is detected, the new task
waits for the older task to complete before it begins execution.

@@

\subsubsection{Isolation}
The reader may wonder why we only compare sibling tasks for
dependences: can't arbitrary pairs of tasks have dependences?  The
answer is ``yes'', but the following lemma shows that
any non-sibling tasks are always detected first as a dependence between
siblings that are registered earlier in execution!
\begin{lemma}
\label{lem:isolation}
\rm
Let $t_1$ and $t_2$ be two distinct, non-sibling tasks other than the root task with parents $p_1$ and $p_2$, respectively.
If $t_1$ and $t_2$ have a dependence, then $p_1$ and $p_2$ have a dependence.
\end{lemma}
\begin{proof}
Recall from Section~\ref{sec:ex} that the parents' region arguments include the children's region arguments and have
only more permissions.
\end{proof}
Lemma~\ref{lem:isolation} justifies checking dependences only between sibling tasks, because any dependences between 
arbitrary tasks are detected first between children of the tasks' least common ancestor.  Legion
guarantees that all dependence analysis is performed for all children of a task $t$ on the instance of the SOOP where $t$
runs, which by Lemma~\ref{lem:isolation} is sufficient to detect all required dependences across all tasks.

@@

>>>>

\item For performance, Legion must hide the extremely long latencies associated
  with machines that have both distributed memory and many levels of
  memory hierarchy.
\end{itemize}
Our scheduler uses three techniques to achieve both correctness and high performance:
\begin{itemize}
\item If it were necessary to compare all pairs of tasks for dependences that 
would be a bottleneck in most Legion programs.  Fortunately,
subtasks enjoy an isolation property that allows Legion to check only task siblings for dependences
(see Section~\ref{sec:dep}).

@@

Each processor in the machine runs an instance of the SOOP, which handles all requests
for Legion services from tasks executing on the processor.  The SOOP is analagous to the out-of-order
instruction scheduler in superscalar processors, but working at the granularity of tasks and their region
arguments instead of instructions and their register arguments.
Legion's SOOP has a five stage pipeline: dependence analysis (Section~\ref{sec:dep}),
distribution (Section~\ref{sec:dist}),
mapping (Section~\ref{sec:map}),
execution (Section~\ref{sec:exec}),
and clean-up (Section~\ref{sec:clean}).
We discuss each stage in turn.

@@

When a Legion task runs, it may spawn new subtasks to be executed.  We refer to the
subtasks as {\em children} of the {\em parent} task.  The children of a parent task
are {\em siblings} of each other.  A task always executes on a single processor, but
its subtasks may execute elsewhere.

When a parent task spawns a child task, the child is {\em registered}
with the SOOP on the processor where the parent is executing;
registration records the subtask's logical region arguments and
associated privileges and coherence properties.  Children are
registered in the sequential execution order in which the parent task
spawns them and enqueued for dependence analysis.  For example, in the
circuit simulation in Listing~\ref{lst:code_ex}, the three spawn
statements on lines 32-34 register will register all of the three
kinds of subtasks (in program order) on the processor where {\tt
simulate\_circuit} executes.

If two children have a dependence between their region arguments, the
SOOP will schedule them to run serially to preserve the program's
sequential semantics.  If two child tasks have no dependence then
Legion can schedule them to execute out of order, in parallel.  Thus,
the first stage in the execution of a task is dependence analysis on
the logical regins used by tasks.

\subsubsection{Dependences}
Detecting dependences between a newly registered task $t$ and a previously registered task 
$t'$ requires comparing the two sets of logical regions accessed.  For each logical region used by
$t'$ that may overlap with a logical region used by $t$, the privileges
and coherence modes are compared to determine whether a dependence exists.  If
both regions need only read privileges there is never a dependence, but if either
task needs write or reduction privileges, the coherence modes are compared using
the table in Figure~\ref{fig:dependence}.

\begin{figure}
{\small
\begin{tabular}{c|cccc}
             & Exclusive & Atomic   & Simultaneous & Relaxed \\
\midrule
Exclusive    & Dep & Dep & Dep & Dep \\ 
Atomic       & Dep & Same & Cont & Cont \\
Simultaneous & Dep & Cont & Same & None \\
Relaxed      & Dep & Cont & None & None \\
\end{tabular}
}
\caption{Dependence table.}
\label{fig:dependence}
\end{figure}
A {\tt Dep} entry indicates dependence while {\tt None}
indicates independence.  A {\tt Same} entry is a dependence unless the two tasks
use the same physical instance of the logical region.\footnote{For example, if two tasks wish to acess the region with atomic coherence, then if they are working on the same physical copy of the data atomicity can be guaranteed using, e.g., locking.  If they are working on different copies of the data then the only safe execution strategy for Legion is to serialize the two tasks.}
A {\tt Cont} entry
indicates a dependence unless a single writer is using the
atomic coherence mode.  When a dependence is detected, the new task
waits for the older task to complete before it begins execution.

@@

\subsubsection{Isolation}

%
%\begin{figure}
%
%\centering
%\begin{tikzpicture}
%\node(top) at (3.5,3.5) { $p$ };
%
%\node(s1) at (1.9,2.5) { $s1$ };
%\node(s2) at (5.1,2.5) { $s2$ };
%
%\node(m1) at (1.9,1.5) { $\ldots$ };
%\node(m2) at (5.1,1.5) { $\ldots$ };
%\node(t1) at (1.9,0.5) { $t_1$ };
%\node(t2) at (5.1,0.5) { $t_2$ };
%
%\draw
%  (top.south) edge (s1.north)
%  (s2.north) edge (top.south)
%  (m1.north) edge (s1.south)
%  (m2.north) edge (s2.south)
%  (t1.north) edge (m1.south)
%  (t2.north) edge (m2.south)
%  ;
%
%\draw[dashed]
%  (t1.east) edge (t2.west)
%  (s1.east) edge (s2.west)
%  ;
%
%\end{tikzpicture}
%\caption{Dependence between tasks implies a dependence between siblings of the least common ancestor.}
%\label{fig:independence}
%\end{figure}

The reader may wonder why we only compare sibling tasks for
dependences: can't arbitrary pairs of tasks have dependences?  The
answer is ``yes'', but the following observation shows why it is sufficient
to check only sibling tasks for dependences.
\begin{observation}
\label{obs:isolation}
\rm
Let $t_1$ and $t_2$ be two sibling tasks with no dependence.  Then no subtask of $t_1$ has a dependence with any subtask
of $t_2$.
\end{observation}
Recall from Section~\ref{sec:ex} that a subtasks can only access subregions of their parent task's region arguments (and with fewer
privileges).  Thus if the regions used by $t_1$ and $t_2$ do not overlap, the regions used by any subtasks of $t_1$ cannot
overlap with the regions used by any subtask of $t_2$.  In the case where the regions used by $t_1$ and $t_2$ do overlap
but there is no dependence because the regions are used with simultaneous or relaxed coherence, then by definition there is no dependence between the subtasks.

@@

Consider the first two subtasks spawned on line 32 of the circuit
simulation in Listing~\ref{lst:code_ex}, {\tt
calc\_new\_currents(pieces[0])} and {\tt
calc\_new\_currents(pieces[1])}.  The first of these tasks reads and
writes the private wires subregion {\tt pieces[0].rw\_pvt} in
exclusive mode, and reads in exclusive mode the node subregions {\tt
pieces[0].rn\_pvt}, {\tt pieces[0].rn\_shr}, and {\tt
pieces[0].rn\_ghost} (see line 39).  The second subtask must be
checked against the first for any dependences; the second subtask uses
{\tt pieces[1].rw\_pvt} (read/write exclusive) and {\tt
pieces[1].rn\_pvt}, {\tt pieces[1].rn\_shr}, and {\tt pieces[1].rn\_ghost} (read-only exclusive).
Recall the region tree in Figure~\ref{sfig:part_fig:tree}.  We reason about the dependences as follows:
\begin{enumerate}
\item {\tt pieces[0].rw\_pvt} and {\tt pieces[1].rw\_pvt} do not overlap as they are subregions of a disjoint partition.
\item {\tt pieces[0].rn\_pvt} and {\tt pieces[1].rn\_pvt} do not overlap for the same reason.
\item {\tt pieces[0].rn\_shr} and {\tt pieces[1].rn\_shr} do not overlap, also for the same reason.
\item {\tt pieces[0].rn\_ghost} may overlap with {\tt pieces[1].rn\_shr}, because they are in two different
  partitions of the same region.
\item {\tt pieces[0].rn\_shr} and {\tt pieces[1].rn\_ghost} may overlap because they are in two different
  partitions of the same region.
\item {\tt pieces[0].rn\_ghost} may overlap with {\tt pieces[1].rn\_ghost} because they are in different subregions of
a non-disjoint, or {\em aliased}, partition.
\item All the remaining pairs of regions used by the two tasks (e.g., {\tt pieces[0].rw\_pvt} and {\tt pieces[1].rn\_pvt})
do not overlap because they are in different region trees and have no common region ancestor.
\end{enumerate}
Cases 4-6 indicate a possible dependence, however in each case the regions are accessed with only read privileges and
so no dependence exists.  Thus, these two subtasks are independent.

A closer look at this example shows that whether there is overlap
between two regions $r_1$ and $r_2$ can be determined by examining
their least common ancestor $\lca{r_1}{r_2}$ in the region tree.  Note
that the region tree alternates between regions and partitions at each
level.  That is, regions are always children of partitions and
partitions are always children of regions in the region tree.  There
are four cases.  If $\lca{r_1}{r_2}$ either does not exist (the regions
are in different region trees) or is a disjoint partition, then $r_1$
and $r_2$ are guaranteed to be disjoint.  If $\lca{r_1}{r_2}$ is an
aliased partition or a region (because $r_1$ and $r_2$ are in
different partitions of the region, as in cases 4 and 5
above), then $r_1$ and $r_2$ may not be disjoint.
